---
title: "Data Preprocessing"
author: "Jiajun Chen"
date: "2025-04-23"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(ggplot2)
library(dplyr)
library(randomForest)
library(tree)
library(lubridate)
library(grid)
library(glmnet)
library(corrplot)
library(tibble)
```
First, we import all the data set we are going to use including the currency conversion
datasets.
```{r}
fhr_data <- read.csv("cleaned_fhr_data.csv")
extract_data <- read.csv("cleaned_detailed_supplier.csv")
conversion_data <- read.csv("conversions.csv")
inflation_data <- read.csv("inflation rate.csv")
```

We add the risk levels that given in extract_data to fhr_data, and convert date columns into
correct format
```{r}
fhr_data<- fhr_data %>% mutate(risk= ifelse(Current.FHR >= 80, "very low risk", ifelse(Current.FHR >= 60, "low risk", ifelse(Current.FHR>= 40, "medium risk", ifelse(Current.FHR>=20, "high risk", "very high risk"))))) 


fhr_data <- fhr_data %>%
  mutate(Period.End.Date = as.Date(Period.End.Date, format = "%m/%d/%Y %H:%M"))
```

Rename the column in extract_data to match the one in fhr_data for merging
```{r}
# convert Vlookup.Supplier.. into Supplier.Number
names(extract_data)[names(extract_data)=="Vlookup.Supplier.."]<- "Supplier.Number"
fhr_static <- fhr_data %>%
  select(Supplier.Number, Data.Source, Group, Group.Classification, Parent.ID)
fhr_static <- fhr_static %>%
  distinct(Supplier.Number, .keep_all = TRUE)
# merge the two datasets that we use for prediction
merged_data <- left_join(extract_data, fhr_static, by = "Supplier.Number")

```
We now want to merge the currency conversion data into the merged financial health data
```{r}
# first merge the conversion_data and inflation_data by year
money_merge <- left_join(conversion_data, inflation_data, by = c("eqyYear" = "Year"))
# make sure all rows with valid conversion values
money_merge <- money_merge %>%
  filter(!is.na(X.Other.Currency.to.USD))
# now merge the 
merged_data <- left_join(merged_data, money_merge, 
                         by = c("eqyYear" = "eqyYear", "currency" = "currency"))
# check the inflation rate for usd and convert other currencies to USD
merged_data <- merged_data %>%
  mutate(Inflation.Rate.... = ifelse(
    currency == "USD",
    inflation_data$Inflation.Rate....[match(eqyYear, inflation_data$Year)],
    Inflation.Rate....
  ))
merged_data <- merged_data %>%
  mutate(X.Other.Currency.to.USD = ifelse(
    currency == "USD",
    1,
    X.Other.Currency.to.USD
  ))

# write to new csv file
write.csv(merged_data, "merged.csv")

```


After we get a data file with USD, we need to convert the monetary values into USD
in 2025 since inflation need to be considered.
## after merged
```{r}
final_merged <- read.csv("final_merged.csv")
inflation_factor<- read.csv("inflation_factor.csv")
# merge the financial health file with final inflation factor
final_merged<-left_join(final_merged, inflation_factor, by = c("eqyYear" = "eqy_year"))
# get non-monetary columns so we exclude them when we convert later
non_monetary <- c("Supplier.Number", "eqyYear", "currency",
                  "X.Other.Currency.to.USD", "inf_factor", "Inflation.Rate....", "FHR", "CHS")
# convert based on inflation factor.
final_merged <- final_merged %>% mutate(across(
    .cols = where(is.numeric) & !non_monetary,
    .fns = ~ . * final_merged$X.Other.Currency.to.USD* final_merged$inf_factor
  ))
# write to the csv file
write.csv(final_merged, "final_merged_updated.csv")
```