---
title: "Capstone2pt2"
author: "Manasa Subramanian.191"
date: "2025-04-01"
output: html_document
---

```{r}
library(broom)
library(tidyr)
library(R.utils)
library(GGally)
library(ggplot2)
library(MASS)
library(dplyr)
library(foreign)
library(tidyverse)
library(readstata13)
library(haven)
library(ISLR)
library(dplyr)
library(corrplot)
library(survival)
library(knitr)
library(readxl)
library(data.table)
library(haven)
library(Amelia)
library(priceR)
library(randomForest)
library(Metrics)
library(dplyr)
library(caret)
library(glmnet)
```


This is random forest to see importance of variables in comparison to the 
```{r}
set.seed(222)
library(dplyr)
train_data <- read_csv("C:/Users/Owner/Downloads/SoccermaticsForPython-master/SoccermaticsForPython-master/Honda_STAT4911_Project/final_previous_merged.csv")
train_data[is.na(train_data)] <- 0
train_data<-train_data %>% select(-c("prev_inf_factor","prev_X.Other.Currency.to.USD","prev_financialDate"))

test_data = read_csv("C:/Users/Owner/Downloads/SoccermaticsForPython-master/SoccermaticsForPython-master/Honda_STAT4911_Project/final_future_merged.csv")

train_cols <- names(train_data)
test_cols <- names(test_data)

# Find the columns that are in 'end' but not in 'most_recent_records'
missing_columns <- setdiff(train_cols, test_cols)

test_data[is.na(test_data)] <- 0

X_train <- train_data[, -which(names(train_data) == "FHR")]  
y_train <- train_data$FHR 

X_test <- test_data[, -which(names(test_data) == "FHR")]  
y_test <- test_data$FHR

rf_model <- randomForest(x = X_train, y = y_train, ntree = 500, importance = TRUE)
y_prediction_rf <- predict(rf_model, newdata = test_data)
mse_rf <- (sum((y_prediction_rf - y_test)^2)/length(y_prediction_rf))
rmse_rf <- sqrt(mse_rf)
print(rmse_rf)
print(paste("RMSE: ",rmse_rf))
imp <- importance(rf_model)
sorted_importance <- sorted_importance <- imp[order(-imp[, 1]), ]
summary(rf_model)
```

```{r}
sorted_df <- as.data.frame(sorted_importance) %>%
  head(10)  # Get top 10 important features

rownames(sorted_df) <- gsub("prev_", "", rownames(sorted_df))

ggplot(sorted_df, aes(x = reorder(rownames(sorted_df), sorted_df[, 1]),
                      y = sorted_df[, 1],
                      fill = sorted_df[, 1])) +
  geom_bar(stat = "identity") +
  coord_flip() +
  theme_minimal() +
  scale_fill_gradient(low = "lightblue", high = "red") +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "none") +
  labs(x = "Features", y = "Importance", title = "Top 10 Random Forest Lag 1 Features")
```
As we can see here the following are of importance in predicting the FHR score:
CHS
debitOwnedWithinOneYear
InterestExpense
titalShareholderEquity
Supplier.Number
RRID
bankCashBalances
Parent.ID
financialAssets
Group
totalCurrentLiabilites
netProfitAfterTax
termLoans
netFundingCashFlow
interestReceived

RIDGE REGRESSION

```{r}
finaldf_ridge <- read_csv("C:/Users/Owner/Downloads/SoccermaticsForPython-master/SoccermaticsForPython-master/Honda_STAT4911_Project/final_previous_merged.csv")

finaldf_ridge<-finaldf_ridge %>% select(-c("prev_inf_factor","prev_X.Other.Currency.to.USD","prev_financialDate"))
```

```{r}
set.seed(222)
split <- createDataPartition(finaldf_ridge$FHR, p = 0.8, list = FALSE)

train_data <- finaldf_ridge[split, ]
train_data[is.na(train_data)] <- 0
test_data <- finaldf_ridge[-split, ]

X_train <- as.matrix(train_data[, -which(names(train_data) == "FHR")])
y_train <- train_data$FHR 
X_test <- as.matrix(test_data[, -which(names(test_data) == "FHR")])
y_test <- test_data$FHR
CV_ridge <- cv.glmnet(X_train,y_train,alpha = 0)
#plotting the lambdas relation to coefficients
plot(CV_ridge)
#goodness of fit
best_lambda <- CV_ridge$lambda.min
y_predicted_ridge <- predict(CV_ridge,newx = X_test, s= "lambda.min")
mse_ridge <- (sum((y_predicted_ridge - y_test)^2)/length(y_predicted_ridge))
print(mse_ridge)
rmse_ridge <- sqrt(mse_ridge)
print(rmse_ridge)
#ridge_coef <- coef(CV_ridge, s = "lambda.min")
#print(ridge_coef)
ridge_coef <- coef(CV_ridge, s = "lambda.min")  # Extract coefficients at best lambda
ridge_coef_df <- as.data.frame(as.matrix(ridge_coef))  # Convert to DataFrame
ridge_coef_df$Feature <- rownames(ridge_coef_df)  # Add feature names
colnames(ridge_coef_df) <- c("Coefficient", "Feature")
ridge_coef_df <- ridge_coef_df[-1, ]  # Remove intercept
ridge_coef_df$Abs_Coefficient <- abs(ridge_coef_df$Coefficient)  # Get absolute values
ridge_coef_df <- ridge_coef_df[order(-ridge_coef_df$Abs_Coefficient), ]

ridge_coef_df$Feature <- gsub("prev_", "", ridge_coef_df$Feature)

top10 <- ridge_coef_df %>% 
  arrange(desc(Abs_Coefficient)) %>% 
  head(10)

ggplot(top10, aes(x = reorder(Feature, Coefficient), y = Coefficient, fill = Coefficient)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  theme_minimal() +
  scale_fill_gradient(low = "lightblue", high = "red") +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "none") +
  labs(x = "Features", y = "Coefficients", title = "Top 10 Ridge Lag 1 Coefficients")

```


```{r}
sorted_df <- as.data.frame(sorted_importance) %>%
  head(10)  # Get top 30 important features

ggplot(sorted_df, aes(x = reorder(rownames(sorted_df), -sorted_df[, 1]), y = sorted_df[, 1])) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  theme_minimal() +
  labs(x = "Features", y = "Importance", title = "Top 10 Features by Importance for Ridge")
```


#Beginning of LASSO

```{r}
library(glmnet)
finaldf <- read.csv("C:/Users/Owner/Downloads/SoccermaticsForPython-master/SoccermaticsForPython-master/Honda_STAT4911_Project/final_previous_merged.csv")
finaldf[is.na(finaldf)] <- 0
for(col in names(finaldf)) {
  if(any(is.character(finaldf[[col]]))) {
    print(paste("Column", col, "contains character values"))
  }
}
```

#LASSO

```{r}

finaldf_lasso<-finaldf %>% select(-c('prev_X.Other.Currency.to.USD','prev_inf_factor','prev_financialDate'))
```

```{r}
sum(is.na(finaldf))
```


```{r}
set.seed(222)
split <- createDataPartition(finaldf_lasso$FHR, p = 0.8, list = FALSE)

train_data <- finaldf_lasso[split, ]
test_data <- finaldf_lasso[-split, ]

X_train <- as.matrix(train_data[, -which(names(train_data) == "FHR")])
y_train <- train_data$FHR 
X_test <- as.matrix(test_data[, -which(names(test_data) == "FHR")])
y_test <- test_data$FHR
CV_lasso <- cv.glmnet(X_train,y_train,alpha = 1)
summary(CV_lasso)
#plotting the lambdas relation to coefficients
plot(CV_lasso)
print(CV_lasso)
#goodness of fit
best_lambda <- CV_lasso$lambda.min
y_predicted_lasso <- predict(CV_lasso,newx = X_test, s= "lambda.min")
mse_lasso <- (sum((y_predicted_lasso - y_test)^2)/length(y_predicted_lasso))
print(mse_lasso)
rmse_lasso <- sqrt(mse_lasso)
print(rmse_lasso)
lasso_coef <- coef(CV_lasso, s = "lambda.min")
print(lasso_coef)
```

```{r}
set.seed(222)
cat("LASSO RMSE:", rmse_lasso, "\n")
cat("Ridge RMSE:", rmse_ridge, "\n")
```

```{r}
lasso_coef_df <- as.data.frame(as.matrix(lasso_coef))  # Convert to DataFrame
lasso_coef_df$Feature <- rownames(lasso_coef_df)  # Add feature names
colnames(lasso_coef_df) <- c("Coefficient", "Feature")
lasso_coef_df <- lasso_coef_df[-1, ]  # Remove intercept
lasso_coef_df$Abs_Coefficient <- abs(lasso_coef_df$Coefficient)  # Get absolute values
lasso_coef_df <- lasso_coef_df[order(-lasso_coef_df$Abs_Coefficient), ]

lasso_coef_df$Feature <- gsub("prev_", "", lasso_coef_df$Feature)

lasso_coef_df <- lasso_coef_df[lasso_coef_df$Abs_Coefficient != 0, ]

top10 <- lasso_coef_df %>% 
  arrange(desc(Abs_Coefficient)) %>% 
  head(10)


ggplot(top10, aes(x = reorder(Feature, Coefficient), y = Coefficient, fill = Coefficient)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  theme_minimal() +
  scale_fill_gradient(low = "pink2", high = "red") +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "none") +
  labs(x = "Features", y = "Coefficients", title = "Top Lasso Lag 1 Coefficients")
```


